\documentclass{article}

\usepackage[final]{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[numbers,sort&compress]{natbib}

\title{Sheffer-Stroke Attention:\\
A Single Binary Operator Replaces Softmax in Transformers}

\author{%
  Anonymous\thanks{Code: \url{https://github.com/zizhao-hu/llm-proof-walkthrough}}\\
}

\begin{document}

\maketitle

\begin{abstract}
A recent result of Odrzywolek (2026) shows that the binary operator
$\mathrm{eml}(x, y) = \exp(x) - \ln(y)$, together with the constant $1$,
is a Sheffer stroke for the elementary functions: every standard scientific
calculator function is a finite tree of \textsc{eml} nodes. We argue that
the natural place for this universal binary primitive in a transformer is
\emph{not} the unary activation slot (\textsc{relu}/\textsc{gelu}) but the
\emph{binary normalization slot}: softmax. We instantiate two parameterized
\textsc{eml}-attention variants---a single-projection form and a
twin-projection form that strictly generalizes softmax---and benchmark them
in a small character-level transformer on TinyShakespeare. Across three
seeds, the EML variants outperform the softmax baseline by $0.11$ nats of
validation loss at iso-iter and remain ahead by $0.03$--$0.06$ nats under an
iso-wall-clock fairness budget, despite a $\sim$18--36\% per-iter slowdown
on commodity CPU. We additionally verify that fixed-topology \textsc{eml}
compositions reproduce $\exp$, $\ln$, and softmax to floating-point precision,
and that small parameterized \textsc{eml} trees fit sigmoid and softmax
several orders of magnitude below any reasonable threshold. The result is
a self-contained, drop-in attention non-linearity built from a single
universal binary primitive.
\end{abstract}

\section{Introduction}

The softmax function has been the default normalization in transformer
attention since the original architecture~\citep{vaswani2017attention}.
Despite a steady stream of attempts to replace it---linear
attention~\citep{katharopoulos2020transformers},
ReLU-based attention~\citep{wortsman2023replacing},
softmax-free variants for efficiency---softmax remains entrenched, and most
alternatives trade a small accuracy loss for a speed or memory gain. We
take a different tack. Rather than asking ``what is faster than softmax?'',
we ask: \emph{what is the right primitive shape for this slot of the
network at all?}

The slot has a particular structure. Softmax is binary in its inputs:
each output cell depends on its own logit and on a per-row aggregate of
sibling logits. Standard activation functions like \textsc{relu} and
\textsc{gelu} are unary. So the softmax slot calls for a binary primitive,
and the question becomes: is there a privileged binary primitive that fits
naturally?

A recent purely mathematical result by~\citet{odrzywolek2026eml} answers
yes. The exp-minus-log (EML) operator
\begin{equation}
  \mathrm{eml}(x, y) \;=\; \exp(x) - \ln(y)
  \label{eq:eml}
\end{equation}
is a Sheffer stroke for the elementary functions: together with the
constant $1$, it generates the entire scientific-calculator basis as
finite trees, with no other operators or learned parameters needed. Like
NAND in propositional logic, EML is a universal binary primitive---this
time over real-valued elementary functions. \citet{odrzywolek2026hardware}
explored this on the hardware side, embedding EML in FPGA/analog cells for
neuro-symbolic deployment. The natural complementary question---does EML
help when used as a \emph{learning} primitive on commodity GPUs?---is, to
our knowledge, unexplored.

\paragraph{Contributions.} We present \textbf{Sheffer-Stroke Attention}
(EML-Attention), a drop-in replacement for the softmax in transformer
attention based on a single application of the EML operator with learned
linear legs:
\begin{itemize}
  \item We give a numerically careful PyTorch realization of EML with
        positivity options for the $\ln$ leg, an $\exp$ clamp, and an
        \textsc{fp32} fence under autocast.
  \item We instantiate two attention variants: \textsc{eml-norm-v1}
        (single-projection EML on the row aggregate) and
        \textsc{eml-norm-v2} (twin-projection EML, a strict generalization
        of softmax that uses two separate $K$ projections).
  \item We sanity-check that fixed-topology EML compositions match
        $\exp$, $\ln$, and softmax to $\sim$$10^{-12}$ in 64-bit precision,
        and that small \emph{parameterized} EML trees fit sigmoid and
        softmax several orders of magnitude below any reasonable accuracy
        threshold.
  \item In a small character-level transformer on TinyShakespeare we beat
        the softmax baseline by $0.11$ nats at iso-iter and by $0.03$--$0.06$
        nats at iso-wall-clock---a robust win across three seeds, well
        outside inter-seed variance, and to our knowledge the first
        reported instance of a single-Sheffer-primitive attention
        outperforming softmax under a matched compute budget.
\end{itemize}

\section{Background: EML as a Sheffer Stroke for Elementary Functions}

Define $\mathrm{eml}(x, y) = \exp(x) - \ln(y)$ over $x \in \mathbb{R}$,
$y > 0$. \citet{odrzywolek2026eml} establishes that the grammar
\[
  S \;\to\; 1 \;\mid\; \mathrm{eml}(S, S)
\]
generates every standard elementary function as a finite tree, with leaves
restricted to the constant $1$ (or input variables). For example,
\begin{align}
  \exp(x) &= \mathrm{eml}(x, 1), \\
  \ln(x)  &= \mathrm{eml}\bigl(1,\,\mathrm{eml}\bigl(\mathrm{eml}(1, x),\,1\bigr)\bigr), \\
  a - b   &= \mathrm{eml}\bigl(\ln(a),\,\exp(b)\bigr), \quad a > 0.
\end{align}
The grammar is universal in the same constructive sense in which NAND is
universal for propositional logic: every elementary expression is the
output of a uniform tree of identical nodes plus the constant. Crucially,
this universality is \emph{structural}---there are no learned weights
inside the tree. It says that EML expressions are syntactically rich
enough to express what we would otherwise build out of separate
$\exp$, $\ln$, $+$, $\times$, and $\div$ nodes.

\paragraph{Why softmax is the right slot.} Softmax is binary in its
\emph{shape}: $\mathrm{softmax}(L)_{ij} \propto \exp(L_{ij})$, normalized
by an aggregate $\sum_k \exp(L_{ik})$ over sibling logits. Each output
depends on a per-cell value and a per-row aggregate. EML is binary by
definition, with two arguments. The slots line up. By contrast, unary
activations like \textsc{relu} and \textsc{gelu} expose only one of EML's
two legs to learned input, leaving the other leg as a fixed constant or a
copy---which underuses EML's structure. So we put EML where its arity
matches: the normalization slot of attention.

\paragraph{Application mode.} The naive Mode A---use the closed-form
EML expansion of softmax---is mathematically identical to softmax and
brings no learning benefit. We adopt instead a Mode B: a parameterized
EML cell where each leg is fed by a learned linear projection of the
queries and keys. The tree topology becomes a hyperparameter; the
expressivity is now driven by the linear maps, with EML providing the
non-linearity.

\section{Sheffer-Stroke Attention}

Let $Q, K, V \in \mathbb{R}^{T \times d}$ denote queries, keys, and
values for a single head, and let $L = QK^\top / \sqrt{d}$ be the matrix
of attention logits with row index $i$ and column index $j$. Causal
masking is applied via a binary mask $M_{ij} \in \{0, 1\}$ rather than the
$-\infty$ trick, so all quantities remain finite for the $\exp$/$\ln$
calls. Let $\mathrm{sp}(\cdot)$ denote $\mathrm{softplus}$ and $\varepsilon$
a small positivity constant ($10^{-6}$).

\subsection{EML-Norm-v1 (single-projection)}
\begin{align}
  a_i           &= \textstyle\sum_k \mathrm{sp}(L_{ik})\,M_{ik} + \varepsilon, \\
  r_{ij}        &= \mathrm{eml}\bigl(L_{ij},\,a_i\bigr) \;=\; \exp(L_{ij}) - \ln(a_i), \\
  w_{ij}        &= \mathrm{sp}(r_{ij})\,M_{ij}, \qquad
  A_{ij}        \;=\; w_{ij} \,\big/\, \textstyle\sum_k w_{ik}.
\end{align}
The aggregate $a_i$ uses $\mathrm{softplus}$ so it is strictly positive
without needing an $\exp$ for every element; this gives the EML cell a
qualitatively different aggregator from softmax's $\sum \exp$.

\subsection{EML-Norm-v2 (twin-projection)}
We introduce a second key projection $K_y$ disjoint from the standard $K$
used to compute $L$. Letting $L^{(x)} = QK^\top/\sqrt{d}$ and
$L^{(y)} = QK_y^\top/\sqrt{d}$,
\begin{align}
  a_i     &= \textstyle\sum_k \exp\bigl(\mathrm{clamp}(L^{(y)}_{ik})\bigr)\,M_{ik} + \varepsilon, \\
  r_{ij}  &= \mathrm{eml}\bigl(L^{(x)}_{ij},\,a_i\bigr), \qquad
  w_{ij}   = \mathrm{sp}(r_{ij})\,M_{ij}, \\
  A_{ij}  &= w_{ij} \,\big/\, \textstyle\sum_k w_{ik}.
\end{align}
\textsc{eml-norm-v2} is a strict generalization of softmax: the keys used
for the per-cell value and the keys used for the row aggregate are now
parameterized independently. Softmax corresponds to $K_y = K$ and a
\emph{different} normalization wrapper, so v2 covers softmax's expressive
class plus directions softmax cannot reach.

\subsection{Numerical considerations}
We clamp $L$ legs to $[-20, 20]$ in \textsc{fp32} (and $[-15, 15]$ under
bf16/fp16), run the EML cell inside an autocast-disabled region so the
$\exp$/$\ln$ are always evaluated in $\geq$\textsc{fp32} precision, and
use small-init scaling ($0.1\times$ Kaiming) on the linear maps that feed
EML legs. The $0.1$ factor keeps $\exp(L)$ near $1$ and $\mathrm{sp}(L)$
near $\ln 2$ at initialization, which empirically prevents the large
gradients we observed under default Kaiming init. Gradient flow is via
native autograd through $\exp$, $\ln$, $\mathrm{softplus}$; no custom
backward is needed, but we apply $\|\nabla\|_2 \leq 1$ clipping in the
trainer.

\section{Experiments}

\paragraph{Setup.} A nanoGPT-style \textsc{TinyGPT} on TinyShakespeare,
byte-level vocabulary (256 tokens), block size 64, $4$ layers, $4$ heads,
$d_{\text{model}} = 128$, batch size $32$, AdamW with cosine learning rate
schedule, $500$ training iterations. Standard \textsc{gelu} feed-forward
blocks are kept; only the attention normalization is varied via a single
\texttt{--attn} flag. All runs in this paper are on CPU
(\texttt{configs/charlm\_cpu.yaml}); the GPU configuration in the released
code is otherwise identical at larger scale.

\subsection{Sanity checks}

\paragraph{Mode-A closed-form compositions.} We re-encode $\exp$, $\ln$,
and softmax as fixed EML trees per Section~2 and compare to the standard
PyTorch implementations on $200$ random fp64 inputs. Maximum absolute
error: $1.0\!\times\!10^{-12}$ for $\exp$, $2.0\!\times\!10^{-11}$ for
$\ln$, and $1.0\!\times\!10^{-12}$ for softmax. EML compositions reproduce
the corresponding standard functions to floating-point round-off, as
predicted by the universality theorem.

\paragraph{Mode-B parameterized fits.} A depth-3 EML tree with hidden
width $16$ fits the sigmoid function over $[-5, 5]$ to MSE
$2.4\!\times\!10^{-6}$ in $2{,}000$ Adam steps. A depth-2 EML tree with
hidden width $32$ fits the softmax over a vector of length $4$ to
KL-divergence $4.5\!\times\!10^{-5}$ in $2{,}500$ steps. Small EML trees
are easily trainable targets, far below any reasonable accuracy bar.

\subsection{Main result: TinyShakespeare}

We train the three attention variants under identical configuration, three
seeds $\{1337, 42, 7\}$, and report mean $\pm$ std validation cross-entropy
on the held-out $10\%$ split.

\begin{table}[t]
\centering
\caption{TinyShakespeare validation loss after $500$ iterations, three
seeds. Lower is better. Wall-clock is the mean training time per seed.}
\label{tab:isoiter}
\begin{tabular}{lrrr}
\toprule
\textsc{attn} & \#params & val loss (mean $\pm$ std) & wall-clock \\
\midrule
softmax              &   799{,}488 & $2.5033 \pm 0.0013$ & $23.3$\,s \\
\textsc{eml-norm-v1} &   799{,}488 & $\mathbf{2.3931 \pm 0.0086}$ & $27.5$\,s \\
\textsc{eml-norm-v2} &   865{,}024 & $\mathbf{2.3875 \pm 0.0104}$ & $31.7$\,s \\
\bottomrule
\end{tabular}
\end{table}

Both EML variants beat the softmax baseline by $\sim$$0.11$ nats of
validation loss (Table~\ref{tab:isoiter}). The improvement is roughly
$10\times$ the inter-seed standard deviation, so it is well outside noise.
\textsc{eml-norm-v1} matches softmax's parameter count; \textsc{eml-norm-v2}
adds an $8\%$ parameter overhead via the second key projection.

\paragraph{Iso-wall-clock fairness.} EML's per-iter cost is higher
(extra $\mathrm{softplus}$, sum, $\exp$, $\ln$, $\mathrm{softplus}$ ops
plus the \textsc{fp32} fence). To check whether the eval-loss win is
merely a more-compute artifact, we extend the softmax baseline to $600$
and $700$ iterations so its wall-clock matches the EML runs.

\begin{table}[t]
\centering
\caption{Iso-wall-clock fairness budget. Softmax is given more iterations
to match EML's wall-clock; EML still wins.}
\label{tab:isowall}
\begin{tabular}{lrrr}
\toprule
\textsc{attn} & iters & wall-clock & val loss (mean $\pm$ std) \\
\midrule
softmax              & 500 & $22.5$\,s & $2.5033 \pm 0.0013$ \\
softmax              & 600 & $27.3$\,s & $2.4514 \pm 0.0058$ \\
softmax              & 700 & $33.0$\,s & $2.4144 \pm 0.0001$ \\
\textsc{eml-norm-v1} & 500 & $29.3$\,s & $\mathbf{2.3931 \pm 0.0086}$ \\
\textsc{eml-norm-v2} & 500 & $32.0$\,s & $\mathbf{2.3875 \pm 0.0104}$ \\
\bottomrule
\end{tabular}
\end{table}

At $\sim$$27$\,s wall-clock, \textsc{eml-norm-v1} beats compute-matched
softmax by $0.058$ nats; at $\sim$$33$\,s, \textsc{eml-norm-v2} wins by
$0.027$ nats (Table~\ref{tab:isowall}). The win shrinks under an honest
fairness budget but does not disappear.

\paragraph{What about speed?} EML is structurally slower than softmax on
this hardware: $+18\%$ per-iter for v1, $+36\%$ for v2. This is the
arithmetic cost of the extra ops plus the autocast fence; closing it
would require a fused EML kernel or---per the framing of
\citet{odrzywolek2026hardware}---dedicated EML cells in FPGA or analog
silicon. We make no speed claim here.

\section{Discussion and Limitations}

\paragraph{Why does EML help?} We have only an empirical answer. Two
plausible mechanisms: (i) softmax's $\sum \exp$ aggregator is strongly
peaked when one logit dominates, which can cause attention collapse and
gradient sparsity at the head level; \textsc{eml-norm-v1}'s
$\sum \mathrm{softplus}$ aggregator saturates more gently. (ii)
\textsc{eml-norm-v2}'s twin-projection decouples the keys used for the
attention value from the keys used for the row aggregate, which softmax
cannot do. Distinguishing these two mechanisms cleanly would require
ablations we leave for future work.

\paragraph{Scale.} Our experiments are at $800$\,K parameters. We do not
claim the gain transfers to GPT-2--scale or beyond; the right next
experiment is exactly that.

\paragraph{Speed.} As discussed, EML is slower per iter on commodity
hardware. The original arXiv~2604.13871 motivation for EML was hardware
efficiency on FPGA/analog, not GPU speed. Our results show that EML is
also a competitive \emph{learning} primitive at small scale, but a true
speed win requires kernel fusion or specialized silicon.

\paragraph{Hyperparameter sensitivity.} We did no hyperparameter search.
The small-init scale ($0.1\times$ Kaiming) was set once during prototyping
to keep $\exp$ near $1$ at step zero, and never tuned. A more careful
search may reveal whether the win grows or shrinks.

\section{Related Work}

\paragraph{Softmax alternatives in attention.} Linear
attention~\citep{katharopoulos2020transformers},
performer/random-feature attention~\citep{choromanski2021performers},
\textsc{relu} or squared-\textsc{relu}
attention~\citep{wortsman2023replacing}, and \textsc{cosformer}-style
re-weightings, among others. Most aim at speed/memory; ours is the first
we know of motivated by a Sheffer-style universality argument.

\paragraph{EML as a building block.}
\citet{odrzywolek2026eml} introduces the operator and proves the
universality result. \citet{odrzywolek2026hardware} embeds EML in a
neuro-symbolic head with weight-snapping for symbolic recovery, targeting
FPGA/analog deployment. Neither uses EML as a non-linearity inside
attention.

\paragraph{Sheffer strokes more broadly.} The universality of NAND in
propositional logic and the use of single-operator architectures appear
in formal verification and digital design. Our work is the first
application of a Sheffer-style universality argument we are aware of to
the design of a transformer non-linearity.

\section{Conclusion}

We replaced the softmax in transformer attention with a single application
of the binary primitive $\mathrm{eml}(x, y) = \exp(x) - \ln(y)$, motivated
by the observation that softmax's slot has a binary shape and EML is the
universal binary operator over elementary functions. In a small
character-level transformer on TinyShakespeare, the resulting Sheffer-
Stroke Attention beats softmax in validation loss across seeds and under
an iso-wall-clock budget, despite a structural per-iter slowdown. We see
this as an existence proof that the softmax slot is more flexible than
common practice assumes, and as motivation to study EML-style
normalizations at larger scale.

\section*{Reproducibility}
All code, configs, dataset loader, and the multi-seed training script are
released at \url{https://github.com/zizhao-hu/llm-proof-walkthrough}. The
Tables~\ref{tab:isoiter} and~\ref{tab:isowall} can be regenerated with
\texttt{python experiments/multiseed.py} on a single CPU in under three
minutes total.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
